---
title: "Clinical Trial Efficacy and Safety Analysis"
subtitle: "CDISC ADaM Pilot 01"
author: "Sushma Gummadi"
date: 28.07.2026

format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    smooth-scroll: true
    code-fold: true
    code-tools: true
    embed-resources: true
    self-contained-math: true
    df-print: paged

execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
library(gt)
library(knitr)

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)

format_column_names <- function(data) {
  names(data) |>
    stringr::str_replace_all("_", " ") |>
    stringr::str_to_title()
}
```

```{r}
#| label: validate-report-inputs
#| include: false

required_outputs <- c(
  here("outputs", "figures", "efficacy_trajectory.png"),
  here("outputs", "tables", "efficacy_treatment_effects.csv"),
  here("outputs", "figures", "top_adverse_events.png"),
  here("outputs", "tables", "overall_safety_summary.csv"),
  here("outputs", "tables", "discontinuation_summary.csv"),
  here("outputs", "tables", "week24_completion_summary.csv")
)

missing_outputs <- required_outputs[
  !file.exists(required_outputs)
]

if (length(missing_outputs) > 0) {
  stop(
    paste(
      "The following required report outputs are missing:",
      paste(missing_outputs, collapse = "\n"),
      sep = "\n"
    )
  )
}
```

```{r}
#| label: import-report-results
#| include: false

efficacy_effects <- readr::read_csv(
  here(
    "outputs",
    "tables",
    "efficacy_treatment_effects.csv"
  ),
  show_col_types = FALSE
)

overall_safety <- readr::read_csv(
  here(
    "outputs",
    "tables",
    "overall_safety_summary.csv"
  ),
  show_col_types = FALSE
)

discontinuation_summary <- readr::read_csv(
  here(
    "outputs",
    "tables",
    "discontinuation_summary.csv"
  ),
  show_col_types = FALSE
)

week24_completion <- readr::read_csv(
  here(
    "outputs",
    "tables",
    "week24_completion_summary.csv"
  ),
  show_col_types = FALSE
)
```

# Executive summary

This report presents a reproducible efficacy and safety analysis using
the publicly available CDISC ADaM Pilot 01 clinical trial data.

The analysis included 254 participants:

- 86 assigned to placebo;
- 84 assigned to Xanomeline High Dose; and
- 84 assigned to Xanomeline Low Dose.

Neither Xanomeline dose demonstrated a statistically significant adjusted
efficacy difference compared with placebo at Week 24.

Treatment-emergent adverse events were more frequent in both Xanomeline
groups than in the placebo group. Study discontinuation was also
substantially higher, while Week 24 completion was lower, in both active
treatment groups.

Overall, the findings suggest an uncertain efficacy signal accompanied
by a less favourable tolerability profile for Xanomeline.

# Background and objectives

The purpose of this project was to apply a reproducible clinical-trial
analytics workflow to publicly available CDISC-standardised data.

The analysis objectives were to:

1. describe the study population;
2. examine ADAS-Cog(11) scores over time;
3. estimate adjusted Week 24 treatment differences;
4. summarise treatment-emergent adverse events;
5. evaluate study discontinuation;
6. evaluate Week 24 completion; and
7. produce a portable clinical-style HTML report.

# Data and methods

## Data source

The analysis used publicly available CDISC ADaM Pilot 01 datasets.

The principal analysis datasets were:

- `ADSL` for participant-level information and treatment assignment;
- `ADQSADAS` for ADAS-Cog(11) efficacy assessments; and
- `ADAE` for adverse-event analyses.

Raw data were imported and processed in earlier project notebooks.
Notebook 5 uses the analysis outputs created by those notebooks.

## Analysis populations

The baseline and efficacy analyses included the 254 randomised
participants represented in the subject-level analysis dataset.

The safety analysis used participants included in the safety population
and their actual treatment assignments.

## Efficacy endpoint

The efficacy endpoint was the ADAS-Cog(11) total score.

Higher ADAS-Cog(11) scores indicate greater cognitive impairment.
Therefore, an increase from baseline represents worsening, while a
negative active-treatment difference relative to placebo numerically
favours Xanomeline.

## Statistical methods

Continuous baseline variables were summarised using medians and
interquartile ranges. Categorical variables were summarised using counts
and percentages.

The adjusted Week 24 efficacy model included:

- treatment group;
- baseline ADAS-Cog(11) score;
- age; and
- sex.

Placebo was used as the reference treatment.

Adjusted treatment effects are presented with standard errors,
95% confidence intervals, and exploratory p-values.

Safety outcomes were summarised as the number and percentage of
participants with at least one event.

# Study population

The randomised population consisted of 254 participants.

Treatment-group sizes were similar:

- Placebo: 86 participants;
- Xanomeline High Dose: 84 participants;
- Xanomeline Low Dose: 84 participants.

The treatment groups were broadly comparable with respect to age, race,
body mass index, education level, and baseline cognitive score. Some
variation was present in sex distribution and median baseline
ADAS-Cog(11) score.

# Efficacy results

## ADAS-Cog(11) trajectory

The following figure presents the unadjusted mean ADAS-Cog(11) score
over time by treatment group.

```{r}
#| label: fig-efficacy-trajectory
#| fig-cap: "Mean ADAS-Cog(11) score over time by treatment group."
#| fig-align: center
#| out-width: "95%"

knitr::include_graphics(
  here(
    "outputs",
    "figures",
    "efficacy_trajectory.png"
  )
)
```

Mean ADAS-Cog(11) scores increased over time in all treatment groups,
indicating worsening cognitive performance.

The raw trajectory should not be interpreted as an adjusted treatment
comparison because treatment groups differed slightly in their baseline
scores.

## Adjusted Week 24 treatment effects

```{r}
#| label: tbl-efficacy-treatment-effects
#| tbl-cap: "Adjusted Week 24 efficacy treatment effects relative to placebo."

efficacy_gt <- efficacy_effects |>
  gt() |>
  cols_label(
    .list = stats::setNames(
      format_column_names(efficacy_effects),
      names(efficacy_effects)
    )
  ) |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 3
  ) |>
  sub_missing(
    missing_text = "Not available"
  ) |>
  tab_header(
    title = md("**Adjusted Efficacy Treatment Effects**"),
    subtitle = "Treatment comparisons relative to placebo"
  ) |>
  tab_options(
    table.width = pct(100),
    heading.align = "center",
    column_labels.font.weight = "bold",
    table.font.size = px(14),
    data_row.padding = px(6)
  )

efficacy_gt
```

For Xanomeline High Dose versus placebo, the adjusted difference was
−1.085 points, with a 95% confidence interval from −2.627 to 0.458 and
a p-value of 0.167.

For Xanomeline Low Dose versus placebo, the adjusted difference was
−0.541 points, with a 95% confidence interval from −2.070 to 0.988 and
a p-value of 0.486.

Both adjusted estimates numerically favoured Xanomeline because they
were negative. However, both confidence intervals included zero, and
neither comparison was statistically significant.

The analysis therefore did not provide conclusive evidence of a Week 24
treatment effect for either Xanomeline dose.

# Safety results

## Common treatment-emergent adverse events

The following figure presents the most frequently reported
treatment-emergent adverse events by treatment group.

```{r}
#| label: fig-top-adverse-events
#| fig-cap: "Most common treatment-emergent adverse events by treatment group."
#| fig-align: center
#| out-width: "95%"

knitr::include_graphics(
  here(
    "outputs",
    "figures",
    "top_adverse_events.png"
  )
)
```

Several dermatological and application-site events were more frequent
in the Xanomeline treatment groups than in the placebo group.

## Overall safety summary

```{r}
#| label: tbl-overall-safety-summary
#| tbl-cap: "Participants with at least one treatment-emergent adverse event."

overall_safety_gt <- overall_safety |>
  gt() |>
  cols_label(
    .list = stats::setNames(
      format_column_names(overall_safety),
      names(overall_safety)
    )
  ) |>
  sub_missing(
    missing_text = "Not available"
  ) |>
  tab_header(
    title = md("**Overall Treatment-Emergent Adverse Event Summary**"),
    subtitle = "Participants with at least one event, n (%)"
  ) |>
  tab_options(
    table.width = pct(100),
    heading.align = "center",
    column_labels.font.weight = "bold",
    table.font.size = px(14),
    data_row.padding = px(6)
  )

overall_safety_gt
```

At least one treatment-emergent adverse event occurred in:

- 75.6% of placebo participants;
- 90.5% of Xanomeline High Dose participants; and
- 91.7% of Xanomeline Low Dose participants.

Possibly or probably treatment-related events occurred in:

- 50.0% of placebo participants;
- 83.3% of Xanomeline High Dose participants; and
- 85.7% of Xanomeline Low Dose participants.

At least one severe treatment-emergent adverse event occurred in:

- 5.8% of placebo participants;
- 9.5% of Xanomeline High Dose participants; and
- 19.0% of Xanomeline Low Dose participants.

These findings indicate a less favourable adverse-event profile for both
Xanomeline treatment groups.

## Study discontinuation

```{r}
#| label: tbl-discontinuation-summary
#| tbl-cap: "Study discontinuation by treatment group."

discontinuation_gt <- discontinuation_summary |>
  gt() |>
  cols_label(
    .list = stats::setNames(
      format_column_names(discontinuation_summary),
      names(discontinuation_summary)
    )
  ) |>
  sub_missing(
    missing_text = "Not available"
  ) |>
  tab_header(
    title = md("**Study Discontinuation Summary**")
  ) |>
  tab_options(
    table.width = pct(75),
    heading.align = "center",
    column_labels.font.weight = "bold",
    table.font.size = px(14),
    data_row.padding = px(6)
  )

discontinuation_gt
```

Study discontinuation occurred in:

- 32.6% of placebo participants;
- 67.9% of Xanomeline High Dose participants; and
- 70.2% of Xanomeline Low Dose participants.

Discontinuation was therefore more than twice as frequent in both
Xanomeline groups as in the placebo group.

The available adverse-event data did not support determining which study
discontinuations were specifically caused by an adverse event because
the adverse-event action variable was missing.

## Week 24 completion

```{r}
#| label: tbl-week24-completion
#| tbl-cap: "Week 24 completion by treatment group."

week24_completion_gt <- week24_completion |>
  gt() |>
  cols_label(
    .list = stats::setNames(
      format_column_names(week24_completion),
      names(week24_completion)
    )
  ) |>
  sub_missing(
    missing_text = "Not available"
  ) |>
  tab_header(
    title = md("**Week 24 Completion Summary**")
  ) |>
  tab_options(
    table.width = pct(75),
    heading.align = "center",
    column_labels.font.weight = "bold",
    table.font.size = px(14),
    data_row.padding = px(6)
  )

week24_completion_gt
```

Week 24 completion was achieved by:

- 69.8% of placebo participants;
- 35.7% of Xanomeline High Dose participants; and
- 33.3% of Xanomeline Low Dose participants.

The lower completion rates in the active-treatment groups correspond
with their substantially higher discontinuation rates.

# Data-quality observations

A review of fatal and serious adverse-event records identified an
apparent source-data inconsistency.

Some adverse-event records with a fatal outcome were not classified as
serious events in the source data.

The source records were retained without modification. The inconsistency
should be considered when interpreting serious and fatal adverse-event
summaries.

The adverse-event action variable was completely missing. Therefore,
the analysis could not determine directly whether individual study
discontinuations were caused by adverse events.

# Validation

The report inputs were validated before the analysis sections were
rendered.

Validation checks included:

- confirmation that all required CSV and PNG outputs existed;
- treatment denominator checks;
- participant-level adverse-event deduplication;
- endpoint and visit checks;
- missing-value checks;
- confidence-interval review; and
- regression diagnostic review.

Formal independent SAS reproduction has not yet been completed.
Therefore, the project should be described as internally validated in R
rather than independently validated in both R and SAS.

# Limitations

This analysis has several limitations.

First, it used a public pilot dataset intended for education, standards
development, and software demonstration. It should not be interpreted as
a regulatory analysis or a definitive reanalysis of the original
clinical trial.

Second, the analyses were exploratory and were not based on a
prespecified regulatory statistical analysis plan.

Third, the Week 24 efficacy model used an analysis of change from
baseline rather than a longitudinal mixed model.

Fourth, no independent missing-data sensitivity analysis was performed.

Fifth, study discontinuation was substantially higher in both
Xanomeline treatment groups. This creates an important interpretive
concern because results for participants who remained in the analysis
may not fully represent participants who discontinued treatment.

Sixth, the adverse-event data contained an apparent inconsistency
between fatal outcome and serious-event classifications.

Finally, formal independent SAS validation remains pending.

# Conclusion

Neither Xanomeline dose demonstrated a statistically significant
adjusted efficacy difference compared with placebo at Week 24.

Both treatment estimates numerically favoured Xanomeline, but their
95% confidence intervals included zero.

Xanomeline-treated participants experienced higher rates of
treatment-emergent adverse events, treatment-related events, severe
events, and study discontinuation than placebo-treated participants.

Both Xanomeline groups also had substantially lower Week 24 completion
rates.

Overall, the results suggest an uncertain efficacy signal accompanied
by a clear tolerability disadvantage for Xanomeline in this exploratory
public-data analysis.

# Reproducibility information

```{r}
#| label: package-information
#| echo: false

package_versions <- tibble::tibble(
  Package = c(
    "R",
    "tidyverse",
    "here",
    "gt",
    "knitr",
    "quarto"
  ),
  Version = c(
    paste(
      R.version$major,
      R.version$minor,
      sep = "."
    ),
    as.character(packageVersion("tidyverse")),
    as.character(packageVersion("here")),
    as.character(packageVersion("gt")),
    as.character(packageVersion("knitr")),
    as.character(packageVersion("quarto"))
  )
)

package_versions |>
  gt::gt() |>
  gt::tab_header(
    title = gt::md("**Software Versions**")
  ) |>
  gt::tab_options(
    table.width = gt::pct(60)
  )
```