---
title: "06_aws_deployment"
format: html
---

---
title: "AWS Deployment"
subtitle: "Deploying the Clinical Trial Report with Amazon S3"
author: "Sushma Gummadi"
date: today

format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true

execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook documents the deployment of the final CDISC ADaM Pilot 01
clinical trial report as a static website using Amazon Simple Storage
Service, or Amazon S3.

The analytical work and final report were created locally using R and
Quarto. The rendered, self-contained HTML report was then deployed to
Amazon S3 for public portfolio access.

# Deployment architecture

The deployment workflow was:

1. Analysis datasets were processed in R.
2. Tables and figures were generated in R.
3. Quarto rendered the final report as a self-contained HTML file.
4. The report was copied and renamed to `index.html`.
5. The file was uploaded to an Amazon S3 bucket.
6. Static website hosting was enabled for the bucket.
7. Public read access was configured for the website object.
8. The S3 website endpoint was tested in a web browser.

# Local report preparation

```{r}
#| label: verify-local-report

library(here)

report_path <- here(
  "reports",
  "r",
  "05_final_report.html"
)

file.exists(report_path)

file.info(report_path)[
  ,
  c(
    "size",
    "mtime"
  )
]
```

# Prepare the deployment directory

```{r}
#| label: prepare-deployment-file

deployment_directory <- here("aws-site")

dir.create(
  deployment_directory,
  showWarnings = FALSE,
  recursive = TRUE
)

deployment_file <- file.path(
  deployment_directory,
  "index.html"
)

file.copy(
  from = report_path,
  to = deployment_file,
  overwrite = TRUE
)

stopifnot(
  file.exists(deployment_file)
)

file.info(deployment_file)[
  ,
  c(
    "size",
    "mtime"
  )
]
```

# AWS service

Amazon S3 was used to host the final static HTML report.

A general-purpose S3 bucket was created for this portfolio project. The
self-contained report was uploaded as `index.html`, and static website
hosting was configured with `index.html` as the index document.

The deployment used only public-read permission for the hosted report.
No AWS access keys, passwords, or confidential data were stored in the
project repository.

# Deployed asset

The deployed website contains the final integrated clinical report,
including:

- study background and objectives;
- analysis methods;
- adjusted efficacy results;
- treatment-emergent adverse-event summaries;
- discontinuation and completion summaries;
- data-quality observations;
- limitations; and
- conclusions.

Because Quarto embedded the figures, styles, and supporting resources in
the final HTML document, only one report file was required for the basic
S3 deployment.

# Validation

Deployment validation included:

- confirming the local report existed;
- confirming the copied `index.html` existed;
- comparing local and deployment file sizes;
- confirming the object was uploaded to S3;
- opening the S3 static website endpoint;
- checking table-of-contents navigation;
- confirming all tables rendered;
- confirming both figures rendered; and
- testing the website in a separate browser window.

# Security considerations

Only public portfolio outputs were deployed.

The following items were not uploaded:

- raw credentials;
- AWS access keys;
- local R session files;
- confidential data;
- project source datasets; and
- files containing personal or sensitive information.

# Limitations

The basic Amazon S3 website endpoint does not provide the same security
and custom-domain features as a production website deployment.

A future enhancement could use Amazon CloudFront to provide HTTPS,
caching, and a secure distribution in front of a private S3 bucket.

# Conclusion

The project demonstrates an end-to-end workflow covering clinical data
analysis, reproducible reporting, and cloud-based publication.

R and Quarto were used to create the analytical outputs and report, while
Amazon S3 was used to deploy the final report as a static website.